%==============================================================================
% Research proposal template (English copy)
%==============================================================================
\documentclass[english]{hogent-article}

% Bibliography file
\addbibresource{proposal.bib}

% Programme information
\studyprogramme{Bachelor of applied information technology}
\course{Bachelor thesis}
\assignmenttype{Research proposal}

\academicyear{2025-2026}

\title{XAI-Supported Evaluation of Model Degradation in Vision and Predictive Maintenance Applications}

\author{Marwan Elkhallouki}
\email{marwan.elkhallouki@student.hogent.be}

\supervisor[Co-promotor]{F. Papadopoulou (Ugent, \href{mailto:Foteini.Papadopoulou@UGent.be}{Foteini.Papadopoulou@UGent.be})}

\specialisation{AI \& Data Engineering}
\keywords{XAI, AI, drift, opencv}

\begin{document}

\begin{abstract}
In the context of significant increased usage of Artificial Intelligence (AI) and Machine Learning (ML) in sectors such as predictive maintenance and computer vision (CV), this paper will focus on the challenge of detecting and handling model degradation by using both Explainable AI (XAI) and Drift-Detection algorithms (DD).
In this paper, we will research XAI techniques such as LIME \autocite{garreau2020explaining}, SHAP \autocite{VandenBroeck2022}, etc. to give a broader image of models experiencing DD.
By combining these techniques with DD algorithms such as MDDM \autocite{Pesaranghader2017}, EDDM \autocite{baena2006early}, etc. we find that the combining of these methods significantly increase the detection and interpretability of model-degradation.
We will use the CMAPSS (NASA) \autocite{Saxena2008} as a case for predictive maintenance to prove this hypotheses as a time based degradation dataset.
The ImageNet-C \autocite{Hendrycks2019} will be used to evaluate if the same model pipeline and techniques can used to evaluate real world examples of model degradation due to faulty sensors or cameras, which will contribute to improving trust in AI-systems used in critical fields.
We expect the results of this paper to make for a increased understanding in model-degradation and drift, and the ways in which to negate these factors in real world application.
\end{abstract}

\tableofcontents

% Main text is in a separate file for convenience
\input{proposal-content}

\printbibliography[heading=bibintoc]

\end{document}
