%---------- Introduction ------------------------------------------------------

\section{Introduction}
\label{sec:introduction}

In the last decade, there has been increased usage of AI and ML in diverse sectors such as predictive maintenance and CV.
This technology provide both boons and challenges, mostly pertaining to the trust of said technology \autocite{Radhakrishnan2020}.
This makes model-degradation a unacceptable in the current day and age for retaining trust and reliability.
Model-degradation refers to the phenomenon by which a AI or ML model degrade in performance due to factors such as data-drift, concept-drift, etc. This can cause systems to behave in unexpected manners and decrease trust in the systems, causing companies to drop these systems for more robust alternatives.
This paper aims to gain a better understanding of how XAI and DD-algorithms can be used in production environments to increase trust and predictability for both users and organizations wishing to use these models.

%---------- State of the art --------------------------------------------------

\section{Literature review}
\label{sec:literature}

The current literature describes a multitude of techniques when talking about handling model-degradation.
This paper will be focusing on two distinct fields namely XAI \autocite{Ribeiro2016} and DD-algorithms \autocite{Bayram2022}.
Both fields will be studied in the context of turbofan motor degradation for predictive maintenance \autocite{Saxena2008}.
Finally these pipelines will be evaluated as a complementary point against simulated covariant shift of images \autocite{Hendrycks2019} to check if the same techniques can be applied for other application not pertaining to predictive maintenance with the use of sensors.

\subsection{Research questions}
The central research question is:
\begin{quote}
  \textit{How can drift-detection and XAI techniques be combined for timely detection and explanation of model degradation in turbofan predictive-maintenance scenarios, and to what extent are these techniques effective at identifying operational drift and image corruption?}
\end{quote}

Supporting questions:
\begin{itemize}
  \item How does model-degradation in turbofan-predictive-maintenance manifest in real world applications?
  \item Which Drift-detection- and Explainable-AI-techniques are most accurate at timely detection of operational drift and model-degradation?
  \item To what extent can the found methods be used when detecting and explaining operational drift, model-degradation and image corruption?
  \item How do the found methods relate to each other in detections speed, accuracy and explainability?
\end{itemize}

\subsection{Datasets}
For this research, the following two datasets will be used: CMAPSS (NASA) \autocite{Saxena2008} and ImageNet-C \autocite{Hendrycks2019}.
CMAPSS (NASA) is a dataset made by NASA simulating gradual failure in turbofan engines, commonly used for testing models in predictive maintenance.
The dataset contains four sub-datasets each showing failure in different manners, severity and speed.
ImageNet-C is a dataset comprised of images and their corrupted counterparts.
The corruptions contained mimic real world image or camera corruptions such as frost, sensor degradation, fog, etc.
These corrupted images will be slowly introduced as to mimic a time-bound degradation.

\subsection{Types of drift}
There are differing types of drift that may occur in real world applications such as Covariant drift, label drift and concept drift \autocite{Bayram2022}.
Covariant drift is a change in input data distribution, whereas label drift is a change in the output labels.
Concept drift is characterized by both a change in input data distribution as well as the output labels.
Each of these types will contribute to the final conclusion and will therefore have to be thoroughly explored before drawing any conclusion.


\subsection{Drift-detection algorithms}
Algorithms such as MDDM \autocite{Pesaranghader2017} and EDDM \autocite{baena2006early} are both designed to detect a change in data-distribution leading to model-degradation. Both are made to give real-time signals when a discrepancy between the aparent results and the expected results are shown to be significant.
More DD-algorithms will be explored in this paper.
\subsection{Drift-detection categories}
There are fundamentally different categories of DD-algorithms that reside under concept-drift \autocite{Bayram2022}:
\begin{itemize}
        \item Data distribution-based detection
        \item Performance-based detection
        \item Multiple hypothesis-based detection
        \item Contextual-based detection
\end{itemize}
This paper will limit it's scope to the first three.

\subsection{XAI techniques}
XAI-techniques like LIME \autocite{garreau2020explaining} and SHAP \autocite{VandenBroeck2022} are designed to give insight into the decision-making-process of AI-models. These techniques can help to identify the cause of model-degradation by analyzing which features contribute most to the predictions.
These have been shown to significantly help with increasing accuracy of domain experts in explaining otherwise difficult to explain models.

%---------- Methodology -------------------------------------------------------

\section{Methodology}
\label{sec:methodology}

\subsection{Model selection}
By using separate datasets for predictive maintenance and image corruption, both will get respectively relevant AI-models:
\begin{itemize}
        \item Predictive maintenance --- Recurrent Neural Networks (RNN)
        \item CV --- Convolutional Neural Networks (CNN)
\end{itemize}

\subsection{Implementation}
We will compare several DD-algorithms (MDDM, EDDM, ADWIN/KSWIN, CUSUM, etc.). Detectors will be applied either on model performance metrics (error rates, RUL error) or on raw/embedded data distributions depending on the method.

\subsection{Integration of XAI techniques}
When a detector flags a window, we will compute XAI attributions (SHAP, LIME, saliency maps) aggregated over pre- and post-drift windows. The comparison highlights which features/time-windows changed importance.

\subsection{Evaluation}
We will evaluate detectors and explanations using the following criteria:
\begin{itemize}
  \item \textbf{Detection latency}: samples/steps until alarm and false positive rate for each detector.
  \item \textbf{Model impact}: accuracy drop on ImageNet-C and MAE/RMSE for RUL on CMAPSS before/after drift.
  \item \textbf{Attribution stability}: shifts in average absolute SHAP/LIME/other values across sliding windows.
  \item \textbf{Computational cost}: runtime/throughput overhead of detectors and XAI methods.
  \item \textbf{PHM/NASA score}: asymmetric penalty for RUL errors as used in CMAPSS evaluations \autocite{Saxena2008}.
\end{itemize}

\subsection{Timeline}
\begin{itemize}
  \item \textbf{Week 1-3}: Literature review and final selection of detectors and XAI methods.
  \item \textbf{Week 3-5}: Implement baseline models for ImageNet-C and CMAPSS.
  \item \textbf{Week 6}: Integrate drift-detection algorithms into the pipelines.
  \item \textbf{Week 7-8}: Integrate XAI methods and run initial evaluations.
  \item \textbf{Week 9-10}: Extensive experiments and analysis.
  \item \textbf{Week 11}: Prepare report and visualizations.
  \item \textbf{Week 12}: Finalize the report and presentation.
\end{itemize}

%---------- Expected results --------------------------------------------------

\section{Expected results and conclusion}
\label{sec:expected_results}

We expect two main outcomes: (1) a quantitative comparison of detectors on ImageNet-C and CMAPSS; and (2) an XAI-driven analysis explaining \textit{why} models degrade. The target audience (ML engineers and reliability teams) should obtain a reproducible pipeline combining detection and diagnosis.

Hypotheses:
\begin{itemize}
  \item Performance- or statistic-based detectors (DDM/ADWIN/KSWIN/CUSUM) will detect drift faster than naive run-to-failure monitoring and will show lower false positive rates on CMAPSS than on ImageNet-C.
  \item XAI measures (SHAP/LIME) will reveal significant shifts in feature/patch importance after drift; these shifts will correlate with performance degradation.
  \item A combined pipeline (detect - explain - decide) will reduce time-to-intervention and unnecessary retraining by producing targeted alerts.
\end{itemize}

Reporting elements:
\begin{itemize}
  \item \textbf{Detection metrics}: latency, false positive rate, true positive rate.
  \item \textbf{Model metrics}: top-1 accuracy drop on ImageNet-C; MAE/RMSE for RUL on CMAPSS before/after drift.
  \item \textbf{XAI metrics}: shift in average absolute SHAP values per feature/sensor and attribution stability.
  \item \textbf{Operational impact}: number of triggers for retrain/human review compared to baseline monitoring.
\end{itemize}

Mock-up visualizations and a reporting table will be produced during the experiments. Example expected table structure:

\begin{table}[h]
  \centering
  \caption{Indicative reporting structure for detector and XAI results}
  \small
  \setlength{\tabcolsep}{4pt}
  \begin{tabular}{p{0.18\linewidth} p{0.26\linewidth} p{0.17\linewidth} p{0.16\linewidth} p{0.19\linewidth}}
      \hline
        \textbf{Dataset} & \textbf{Detector/XAI} & \textbf{Latency $\downarrow$} & \textbf{FP-rate $\downarrow$} & \textbf{Attribution shift} \\
      \hline
      ImageNet-C & DDM / ADWIN / KSWIN / CUSUM & \textit{n} samples & \textit{x\%} & Patch-importance shifts \\
      ImageNet-C & SHAP / LIME & n/a & n/a & Top-k patches differ post-drift \\
      CMAPSS & DDM / ADWIN / KSWIN / CUSUM & \textit{n} cycles & \textit{x\%} & Sensor importance shifts \\
      CMAPSS & SHAP / LIME & n/a & n/a & Top sensors differ post-drift \\
      \hline
    \end{tabular}
\end{table}

Value for practitioners:
\begin{itemize}
  \item A reproducible pipeline (detection + XAI) that provides faster, more focused alerts than simple performance monitoring.
  \item Clear visualizations for operators that explain which features or image patches changed and why this points to degradation.
  \item Practical guidelines on when to retrain, when to involve humans, and which detector suits which context.
\end{itemize}

