%%=============================================================================
%% Samenvatting
%%=============================================================================

\chapter*{Abstract}

The gradual decline in model performance caused by shifts in real-world data, what the literature calls model degradation, has become a pressing concern for organisations that rely on AI systems in production. Industrial visual inspection is one of the more vulnerable domains. Cameras age, lighting conditions change, and entirely new defect types can appear that the model has simply never encountered. These shifts rarely announce themselves cleanly, which is exactly what makes them difficult to manage.

This thesis investigates whether drift-detection algorithms and Explainable AI (XAI) techniques can be combined effectively, not just to catch degradation early, but to explain what changed and why. The main dataset is the MVTec Anomaly Detection Dataset, a widely used benchmark for industrial visual inspection covering fifteen object and texture categories. ImageNet-C serves as a secondary validation set, simulating the kind of incremental camera and sensor degradation that happens in practice through a controlled suite of image corruptions across five severity levels.

The experimental work is structured in three phases. A ResNet-50 is first fine-tuned as a binary defect classifier on MVTec AD. Then four drift detectors, DDM, EDDM, ADWIN, and MDDM, are woven into the inference pipeline to track shifts in the error distribution over sliding windows. Finally, XAI methods (LIME, SHAP, and Grad-CAM) are applied to windows around detected drift events to identify which visual features actually changed.

The results are fairly clear on a few points. ADWIN achieves the shortest detection latency, while EDDM produces the fewest false positives. Of the three XAI methods, Grad-CAM is the most immediately interpretable for visualising how the model's attention shifts, whereas SHAP offers a more quantifiable account of attribution change. Together, the combined pipeline substantially reduces time-to-intervention compared to a naive accuracy-threshold monitor, and the findings hold up on the ImageNet-C corruption scenarios as well.

The main contribution here is a reproducible detection-and-explanation pipeline, along with practical guidelines aimed at ML engineers and reliability teams working with visual inspection systems in production settings.

